A Type System for String Sanitation\\Implemented Inside a Python

### 2014-04-09 sanitation-plas14/sanitation-plas14.tex
Added PLAS14 template

\title{\Ace: Active Typechecking and Translation Inside a Python}

\authorinfo{~}{~}{~}
%\authorinfo{Cyrus Omar\and Nathan Fulton\and Jonathan Aldrich}
 %          {School of Computer Science\\
  %          Carnegie Mellon University}
   %        {\{comar, nfulton, aldrich\}@cs.cmu.edu}

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
%gain access to new abstractions, even when they agree that these abstractions
%could be valuable. This suggests a need for languages that satisfy two key
%growth criteria: compatibility with existing programming ecosystems; and
%internal extensibility, so that new abstractions do not require new
%languages.
%
%We introduce Ace, a language compatible with the popular Python ecosystem.
%While Python, like other similar languages, is satisfactory for simple
%scripting, Ace is designed for more complex situations where static
%typechecking is beneficial. Unlike most statically-typed languages, Ace's type
%system can be extended from within, by associating compile-time functions with
%type definitions. The compiler selectively invokes these according to a
%type-directed protocol, avoiding interference problems. Despite these design
%constraints, active types in Ace can express the static and dynamic semantics
%of a range of functional, object-oriented, parallel and domain-specific
%abstractions, all within libraries.
Programmers are justifiably reluctant to adopt new language dialects to access stronger type systems. This suggests a need for a language that is \emph{compatible} with existing libraries, tools and infrastructure and that has an \emph{internally extensible type system}, so that adopting and combining type systems requires only importing libraries in the usual way, without the possibility of link-time ambiguities or safety issues.

We introduce Ace, an extensible statically typed language embedded within and compatible with Python, a widely-adopted dynamically typed language. Python serves as Ace's type-level language. Rather than building in a particular set of type constructors, Ace introduces a novel extension mechanism, \emph{active typechecking and translation}, organized around a bidirectional type system that inverts control over typechecking and translation to user-defined type constructors according to a protocol that cannot cause ambiguities and type safety issues at link-time. In addition to describing a full-scale language design, we give a simplified calculus that describes the foundations of this mechanism and show that, as implemented in Ace, it is flexible enough to admit practical library-based implementations of types drawn from functional, object-oriented, parallel and domain-specific languages. 
%We also show how to construct safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into Ace and then through to OpenCL. % can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library. %We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular. %Ace can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces Ace, a language that aims to satisfy this more comprehensive set of design criteria. To do so, Ace introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of Ace is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}\label{intro}
Asking programmers to import a new library is simpler than asking them to adopt a new programming language. Indeed, recent empirical studies underscore the difficulties of driving languages into adoption, finding that extrinsic factors like compatibility with large existing code bases,  library support, team familiarity and tool support are at least as important as intrinsic features of the language  \cite{Meyerovich:2013:EAP:2509136.2509515,chen05,nguyen2010survey}.

Unfortunately, researchers and domain experts aiming to provide potentially useful new abstractions can sometimes find it difficult to implement them as libraries, particularly when they require strengthening a language's type system. In these situations, abstraction providers often develop a new language or language dialect. Unfortunately, this \emph{language-oriented approach} \cite{journals/stp/Ward94} does not scale to large applications: using  components written in languages with different type systems can be awkward and lead to safety problems and performance overhead at interface boundaries. %To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec. \ref{related-work}).

This is a problem even when a dialect introduces only a small number of new constructs. For example, a recent  study \cite{cave2010comparing} comparing a Java dialect, Habanero-Java (HJ), with a comparable library, \verb|java.util.concurrent|, found that the language-based abstractions in HJ were easier to use and provided useful static guarantees. Nevertheless, it concluded that the library-based abstractions remained more practical outside the classroom because HJ, as a distinct dialect of Java with its own type system, would be difficult to use in settings where some developers had not adopted it, but needed to interface with code that had adopted it. It would also be difficult to use it in combination with other abstractions also implemented as dialects of Java. Moreover, its tool support is more limited. 
This suggests that today, programmers and development teams cannot use  the abstractions they might prefer because they are only available bundled with languages they cannot adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}.

 %This issue was perhaps most succinctly expressed by a participant in a recent study by Basili et al. \cite{basili2008understanding} who stated ``I hate MPI, I hate C++. [But] if I had to choose again, I would probably choose the same.''
%Unfortunately, researchers and domain experts (collectively, \emph{providers}) who design potentially useful new abstractions often find it difficult to implement them in terms of the general-purpose abstraction mechanisms available in existing languages. 
%Using a foreign construct can be  unnatural and, if the languages have  different type systems, can also cause safety problems.

Internally extensible languages promise to reduce the need for new dialects by giving abstraction providers more direct control over a language's syntax and static semantics from within libraries. %In   an internally extensible language, programmers  can \emph{granularly} import the primitive constructs best suited to each component of their application or library. Providers can thus more easily develop, deploy and evaluate new abstractions in the context of existing codebases, narrowing one of the gaps between research and practice \cite{basili2008understanding}. 
%Indeed, abstractions could compete on their individual merit, rather than  by the serious adoption barriers mentioned above.
Unfortunately, the mechanisms available today have several problems. First, they are themselves often available only within a dialect of an existing language and thus face a ``chicken-and-egg'' problem: a language like SugarJ \cite{erdweg2011sugarj} must overcome the same extrinsic issues as a language like HJ. Second, giving abstraction providers too much control over the language (by introducing an over-zealous ``solution'' to the \emph{expression problem}) can introduce type safety issues and ambiguities that only become apparent when extensions are combined, as we will discuss. Too little control, on the other hand, leaves it difficult to implement real-world abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined. %This requires further research before it would be appropriate to widely rely on them (which we will discuss as we go on).

In this paper, we describe an extensible statically-typed language, Ace, that is implemented entirely as a library within Python and flexibly repurposes its syntax to enable, for example, type annotations. This avoids the ``chicken-and-egg problem'' and occupies what we consider a ``sweet spot'' in a design space that includes the extrinsic criteria described above. We justify this by showing a variety of non-trivial type system fragments that can be embedded as composable libraries within Ace (contribution 1, Sec. \ref{usage}).

Ace, together with a core lambda calculus, serves also as a vehicle for a more general contribution: a novel  underlying extension mechanism, which we call \emph{active typechecking and translation}. As suggested by the design of Ace, it sidestep the expression problem by leaving the forms of expressions fixed. Instead, abstraction providers extend the language by introducing new type constructors and operators using type-level functions. We show how, by  structuring the language as a bidirectional type system and integrating techniques from the typed compilation literature directly into the language, we can maintain expressiveness while  guaranteeing that link-time ambiguities cannot occur and type safety is maintained (contribution 2, Secs. \ref{att} and \ref{theory}).

%Ace can be used from the shell, producing source files that can be further compiled or executed by external means. Ace functions targeting a language with Python bindings (e.g. Python itself, as well as the others just described) can also be compiled and invoked interactively from Python. 
%Ace supports a form of staged \emph{ad hoc} polymorphism (based on singleton types) wherein the ``dynamic type'' of a Python value can determine a static type and trigger active typechecking and translation ``just-in-time'' when Python code calls a generic Ace function. We show how this can significantly streamline interactive workflows (e.g. in scientific computing) that use Python for orchestration and exploratory tasks (e.g. plotting) but call into a statically-typed  language via a foreign-function interface (FFI) for performance- and safety-critical portions. We provide a complete implementation of the OpenCL type system (a language for low-level data parallel programming on GPUs), show several higher-level extensions to it, and discuss bindings that bridge between the type systems of Ace, OpenCL and Python's standard numerics package, \verb|numpy|, all as a library (contribution 3, Sec. \ref{targets}).%We will discuss in detail as our primary realistic case study.

%The remainder of the paper is organized as follows: in Sec. \ref{usage}, we describe the basic structure and usage of Ace with an example that uses several interesting Ace type libraries. In Sec. \ref{att}, we detail how active typing and translation delegate control, at compile-time, to active types and bases, and discuss how this represents a form of extensibility that side-steps the expression problem. In Sec. \ref{theory}, we reduce this mechanism to  a core lambda calculus, $\lamAce$,  clarify the relationship to   type-level computation and show how type safety can be maintained. In Sec. \ref{targets}, we discuss generic functions, FFIs and staging via a complete implementation of the OpenCL type system (and extensions to it) as a library.

In Sec.  \ref{related}, we place Ace in the context of related work on language extensibility, term rewriting, staged  compilation, bidirectional typechecking, type-level computation and typed compilation. We conclude in Sec. \ref{discussion} by summarizing the key features needed by a host language to support active typechecking and translation, and discussing present limitations and potential future work. %The appendix has details omitted for concision as well as several more examples, including functional datatypes, flexible complex numbers and a previously little-explored type system for tracking string invariants using regular expressions statically (which we show being used in Listing \ref{example}, amongst  other examples).

\section{Language Design and Usage}\label{usage}

%
Listing \ref{example} shows that the top-level of every Ace file is a \emph{compilation script} written directly in Python. Ace requires no modifications to the language (version 2.6+ or 3.0+) or features specific to the primary implementation, CPython (so Ace supports alternative implementations like Jython and PyPy). This choice pays immediate dividends on the first five lines: Ace's import mechanism is Python's import mechanism, so Python's build tools (e.g. \verb|pip|) and package repostories (e.g. \verb|PyPI|) are directly available for distributing Ace-based libraries, including those defining type systems. %Ace can be installed with a single command: \emph{omitted for review}. %Ace is agnostic about the particular design choices that we made in our examples.

%The top-level statements in an Ace file, like the \verb|print| statement on line 10, are executed to control the compile-time behavior, rather than the run-time behavior, of the program. %That is, Python serves as the \emph{compile-time metalanguage} (and, as we will see shortly, the \emph{type-level language}) of Ace. %For readers familiar with C/C++, Python can be thought of as serving a role similar to (but more general than) its preprocessor and template system (as we will see).

\subsection{Types}
\begin{codelisting}
\begin{lstlisting}
from examples.py import py, string
from examples.fp import record
from examples.oo import proto
from examples.num import decimal
from examples.regex import string_in

print "Hello, compile-time world!"

A = record['amount' : decimal[2]]
C = record[
  'name' : string, 
  'account_num' : string_in[r'\d{10}'],
  'routing_num' : string_in[r'\d{2}-\d{4}/\d{4}']]
Transfer = proto[A, C]

@py
def log_transfer(t):
  """Logs a transfer to the console."""
  {t : Transfer}
  print "Transferring %s to %s." % (
    [string](t.amount), t.name)

@py
def __toplevel__():
  print "Hello, run-time world!"
  common = {name: "Annie Ace", 
            account_num: "0000000001", 
            routing_num: "00-0000/0001"} (C)
  t1 = ({amount: 5.50}, common) (Transfer)
  t2 = ({amount: 15.00}, common) (Transfer)
  log_transfer(t1)
  log_transfer(t2)

print "Goodbye, compile-time world!"
\end{lstlisting}
\caption{[\texttt{listing\ref{example}.py}] An Ace compilation script.}
\label{example}
\end{codelisting}

Types are constructed programmatically during execution of the compilation script. This stands in contrast to many  contemporary statically-typed languages, where types (e.g. datatypes, classes, structs) can only be declared. Put another way, Ace supports \emph{type-level computation} and Python is its type-level language (discussed further in Sec. \ref{theory} and Sec. \ref{related}). 
%The compilation script can assign shorter names to types for convenience (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell). 
In our example, we see several types being constructed:
\begin{enumerate}
\item On line 9, we construct a functional record type with a single (immutable) field named \verb|amount| with type \verb|decimal[2]|, which classifies decimal numbers with two decimal places. \verb|record| and \verb|decimal| are \emph{indexed type constructors}. We will provide more details in Sec. \ref{att}, but note that syntactically, \verb|record| overloads subscripting and borrows Python's syntax for array slices (e.g. \verb|a[min:max]|) to approximate conventional functional notation for type annotations. Note also that the field name could equivalently have been written \lstinline{'a' + 'mount'}, again emphasizing that types and indices are constructed programatically by a Python script.
\item On lines 10-13, we construct another record type. The field \verb|name| has type \verb|string|, defined in \verb|examples.py|, while the fields \verb|account_num| and \verb|routing_num| have more interesting types, classifying strings guaranteed statically to be in a regular language specified statically by a  regular expression pattern (written, to avoid needing to escape backslashes, using Python's \emph{raw string literals}). The appendix will give the typechecking rules, based on \cite{fulton-thesis}. To our knowledge, this type system has not previously been implemented. We include it to emphasize that we aim to support specialized type systems,  not just general purpose constructs like records and strings.
\item 
%Prototypic inheritance  is a feature of some dynamically-typed languages (most notably, Self \cite{Ungar:Smith:oopsla:1987} and Javascript). 
On line 14, we construct a simple \emph{prototypic object type} \cite{Lie86}. The type \verb|Transfer| classifies terms consisting of a \emph{fore} of type \verb|A| and a \emph{prototype} of type \verb|C|. If a field cannot be found in the fore, the type system will delegate (here, statically) to the prototype. This makes it easy to share the values of the fields of a common prototype amongst many fores, here to allow us to describe multiple transfers differing only in amount.
\end{enumerate}
\subsection{Typed Functions}

\emph{Typed functions} implement the run-time behavior and are distinguished by the presence of a decorator specifying their  \emph{base semantics} (here \verb|py| from \verb|examples.py|; Sec. \ref{att}). These functions are, however, still written using Python's syntax, a choice that is again valuable for extrinsic reasons: users of Ace can use a variety of  tools designed to work with Python source code  without  modification, including code highlighters (like the one used in generating this paper), editor plugins, style checkers and documentation generators. We will see several examples of how, with a bit of cleverness, Python's syntax can be repurposed within these functions to support a variety of static and dynamic semantics that differ from Python's. Ace  leverages the \verb|inspect| and \verb|ast| modules in the Python standard library to extract abstract syntax trees for functions annotated in this way \cite{python}.

On lines 16-21, we see the function \verb|log_transfer|. We follow Python conventions by starting with a documentation string  that informally specifies the behavior of the function. Before moving into the body, however, we  also write a \emph{type signature} (line 17) stating that the type of \verb|t| is \verb|Transfer|, the prototypic object type described above. As a result, we can assume on line 19 that \verb|t.amount| and \verb|t.name| have types \verb|decimal[2]| and \verb|string|, respectively. We will return to the details of \verb|print| and the form \verb|[string](t.amount)|, which performs an explicit conversion, in Sec. \ref{att}.

Line 19 repurposes Python's syntax for dictionary literals to approximate  conventional notation for type annotations.
%\footnote{In the rare situation where there are no type annotations and the first statement is a dictionary literal, users can use Python's \texttt{pass} keyword.} % (and following conventions for documentation strings, which also describe ``types'' at the end). % The ``keyword'' \verb|type| was chosen because it is the name of a built-in function in Python (and thus would not be used in the way shown for any other purpose).
 {In version 3.0 of Python, syntax for annotating arguments with arbitrary values directly was introduced \cite{pep3107}:
% \vspace{-3px}
\begin{lstlisting}[numbers=none]
def log_transfer(t : Transfer): 
  print ...
\end{lstlisting}
\vspace{-5px}
These annotations were initially intended to serve only as documentation (suggesting that users of dynamically typed languages see potential in a typing discipline, if not a static one), but they were also made available as metadata for use by unspecified  future libraries. Ace supports both notations when the compilation script is run using Python 3.\textit{x}, but we use the more universally available notation here in view of our extrinsic goals: the Python 2.\text{x} series  remains the most widely used by a large margin as of this writing.% (though we consider cross-compilation in Sec. \ref{targets}).

\subsection{Bidirectional Typechecking of Introductory Forms}
On lines 23-32, we define the function \verb|__toplevel__| (the base will consider this a special name for the purposes of compilation, discussed in Sec. \ref{compilation}). After printing a run-time greeting, we introduce a value of type \verb|C| on lines 26-28. Recall that \verb|C| is a record type, so we provide the names of the fields (here, without quotes because we are no longer in the type-level language) and values of the appropriate type using the syntactic form normally used for dictionary literals. We specify which record type the literal should be {analyzed} against by giving a \emph{literal ascription}, \verb|(C)|. This again repurposes existing  syntax: here function application. When the ``function'' is of literal form (dictionaries, tuples, lists, strings, numbers, booleans and  \verb|None|) and the argument is a type, we will treat it instead an \emph{ascribed introductory form} of that type. We see another such form used with Python's tuple syntax on lines 29-30 to introduce terms of type \verb|Transfer| by providing the fore and prototype.

%\subsection{Bidirectional Typechecking}
The string, number and dictionary literal forms inside the outermost forms on lines 26-30 are also  introductory, but do not have an ascription. This is because the outermost ascriptions completely determine which types they will need to have. As we will discuss in more detail shortly, Ace is built around a \emph{bidirectional type system} that distinguishes locations where an expression must \emph{synthesize} a type (e.g. to the right of bindings) from those where it can be \emph{analyzed} against a known type (e.g. as an argument to a function with known type) \cite{Lovas08abidirectional}. Unascribed literal forms must ultimately be analyzed against a type.

If an unascribed literal is used in a synthetic location, the base can, however, specify a ``default ascription''. For example, if we had not specified the literal ascription on line 26, the base would cause the dictionary literal to be analyzed against the type \verb|dyn|, covering dynamically classified Python values. This would still lead to a type error because keys are then treated as expressions, as in Python, rather than field names, and the identifiers shown have not been bound in the top-level context (emphasizing that ``dynamically-typed languages'' can still have a static semantics, even if it only involves checking that top-level variables are bound).% When considered as a record, the keys were considered in a different class by the implementation (labels, rather than variables).

An ascription can also be a {type constructor} instead of a type. For example, we might write \verb|[1, 2] (matrix)| instead of \verb|[1, 2] (matrix[i32])| when using a base for which the default ascription for number literals is \verb|i32|. Because the arguments synthesize the appropriate type, the type constructor can synthesize an appropriate index based on the types of the subexpressions. If we wanted a matrix of \verb|dyn|s, then we would have to write either \verb|[1, 2] (matrix[f32])|, ascribe each inner literal so that it synthesizes the \verb|f32| type, \verb|[1(f32), 2(f32)] (matrix)|, or construct a base with a different default ascription.

We will see how bidirectional typechecking  works in greater technical detail in Secs. \ref{att} and \ref{theory}, but note here that having the semantics of a form change depending on the type it is being analyzed against is a key to achieving expressiveness given our constraint that we cannot add  new forms to the language, for both extrinsic and intrinsic reasons.

\subsection{External Compilation}\label{compilation} 
\begin{codelisting}
\begin{lstlisting}[style=Bash]
$ `acec listing1.py`
Hello, compile-time world!
Goodbye, compile-time world!
[acec] _listing1.py successfully generated.
$ `python _listing1.py`
Hello, run-time world!
Transferring 5.50 to Annie Ace.
Transferring 15.00 to Annie Ace.
\end{lstlisting}
\caption{Compiling \texttt{listing\ref{example}.py} using \texttt{acec}. Both steps can be performed at once by writing \lstinline[style=Bash]{`ace listing1.py`} (line 3 will not be printed with this command).}
\label{external-compilation}
\end{codelisting}
\begin{codelisting}[t]
\begin{lstlisting}
import examples.num.runtime as __ace_0

def print_transfer(t):
  print ("Transferring %s to %s." % (
    __ace_0.decimal_to_str(t[0],2), t[1][0]))

print "Hello, run-time world!"
common = ("Annie Ace", "0000000001", "00-0000/0000")
t1 = ((5,50), common)
t2 = ((15,0), common)
log_transfer(t1)
log_transfer(t2)
\end{lstlisting}
\caption{[\texttt{\_listing\ref{example}.py}] The file generated in Listing \ref{external-compilation}.}
\label{example-out}
\end{codelisting}

To typecheck, translate and execute the code in Listing \ref{example}, we have two choices: do so externally at the shell, or perform compilation interactively (or implicitly) from within Python. We will begin with the former. Interactive and implicit compilation are implemented but we do lack space to detail it in the present paper, choosing here to focus on the core mechanism.% with our OpenCL examples.

Listing \ref{external-compilation} shows how to invoke the \verb|acec| compiler at the shell to typecheck and translate \texttt{listing\ref{example}.py}, resulting in a file named \verb|_listing1.py|. This is then sent to the  Python interpreter for execution. Note that the {\texttt{print}} statements at the top-level of the  compilation script were evaluated during compilation only. These two steps can be combined by running \verb|ace listing1.py| (the intermediate file is not generated unless explicitly requested in this case).

The Ace compiler is itself a Python library, and \verb|acec| is a simple Python script that invokes it, operating in two steps:
\begin{enumerate}
\item It evaluates the compilation script to completion.
\item For any top-level bindings that are Ace functions in the final environment (instances of \verb|ace.TypedFn|, as we will discuss), it initiates active type-checking and translation (Sec. \ref{att}). If no type errors are discovered, the translations are collected (obeying order dependencies) and emitted, with file extension(s) determined by the target(s) in use, discussed further in Sec. \ref{att}. Here our target is  the default target associated with the base \verb|py|, which emits Python 2.6 files. If a type error is discovered, no file is emitted and the error is displayed on the console.
\end{enumerate}

In our example, there are no type errors, so the file \verb|_listing1.py|, shown in Listing \ref{example-out},  is generated. This file is meant only to be executed. The invariants necessary to ensure that execution does not ``go wrong'' were checked statically and entities having no bearing on execution, like field names and types themselves, were erased. Notice that:
\begin{enumerate}
\item The base recognized the function name \verb|__toplevel__| as special, placing the translation of its body  at the top level of the file. Ace does not have any special function names itself; this is a feature of the user-defined base.
\item Records with two or more fields translated to tuples of their values (e.g. \verb|common| on line 8). If there was only a single field, like the terms of type \verb|A| inside \verb|t1| and \verb|t2|, the value was passed around unadorned. 
\item Decimals translated to pairs of integers. Conversion to a string happened via a helper function defined in a ``runtime'' package imported with an internal name, \verb|__ace_0|, to avoid naming conflicts. 
\item Terms of type \verb|string_in[r"..."]| translated to strings. Checks for membership in the specified regular language were performed entirely statically by the type system.\footnote{Note that there are situations (e.g. if a string is read in from the console) that would necessitate an initial run-time check, but no further checks in downstream functions would be necessary. See the appendix for details.}
\item Prototypic objects are represented as pairs consisting of the fore and the prototype. Dispatch to the appropriate record based on the field name is static (line 5).
\end{enumerate}

% As a result, execution is essentially as fast as it could be given the target language chosen (Python). We will target an even faster language, \verb|OpenCL|, in Sec. \ref{targets}.

%Because \verb|t| will be statically guaranteed to have type \verb|Transfer|,  
\subsection{Type Errors}
\begin{codelisting}[t]
\begin{lstlisting}
from listing1 import py, A, C, log_transfer
from datetime import date
@py
def pay_oopsie(a):
  {a : A}
  print "Checking date..."
  if date.today().day == 1:
    common = {nome: "Oopsie Daisy",
              account_num: None,
              routing_num: "0-0000-0002"} (C)
    log_transfer((common, a))
    a.amount += 1000 
print str(date.today().day)
\end{lstlisting}
\caption{[\texttt{listing\ref{oops}.py}] Lines 9-13 each have type errors.}
\label{oops}
\end{codelisting}
\begin{codelisting}
\begin{lstlisting}[style=Bash]
$ `acec listing4.py`
2
[acec] TypeError in listing4.py (line 8, col 15): 
  [record] Invalid field name: nome
           Expected: name, account_num, routing_num
\end{lstlisting}
\caption{Compiling \texttt{listing\ref{oops}.py} using \texttt{acec} catches the errors statically (compilation stops at first error).}
\label{oops-compilation}
\end{codelisting}Listing \ref{oops} shows an example of code containing several type errors. Indeed, lines 8-12 each contain a type error. If analagous code were written in Python itself, these could only be found if the code was executed on the first day of the month (and, depending on the implementation, not all of the issues would  immediately result in run-time exceptions, possibly leading to quite subtle problems; for example, the unexpected use of \verb|None|\footnote{The notation \texttt{+T} constructs a \texttt{None}-able option type; see appendix.}). Static type checking allows us to find these errors during compilation. Listing \ref{oops-compilation} shows the result of attempting to compile this code. The compilation script completes (so that functions can refer to each other mutually recursively), then the typed functions in the top-level environment are typechecked. The typechecker raises an exception statically at the first error, and \verb|acec| prints it to the console as shown.

\section{Active Typechecking and Translation}\label{att}

%With monolithic languages, enabling a variety of tools is relatively straightforward. For example, one might declare functional datatypes with cases for every type constructor and term form, and then proceed by  exhaustive case analysis over term forms to implement this sort of functionality. The external visitor pattern for class-based object systems operates in essentially the same way. In both cases, the problem is that the types and terms are defined in one place, so  providers cannot extend the language modularly.

Enabling the addition of new forms to a language in an open tool ecosystem is difficult. Indeed, it is the canonical example of the long-studied \emph{expression problem} \cite{wadler1998expression}. Although a number of approaches have been developed, we argue that many of them are overly permissive, making it difficult to reason compositionally about metatheoretic issues (e.g. type safety) and avoid ambiguities when extensions are combined (see Sec. \ref{related}). %They are also often quite complex. %Requiring that abstraction clients have the expertise to resolve these ambiguities manually is undesirable, and heading them off results in an unbounded, combinatoric explosion of situations to consider.
Ace's term forms, as we have seen, are fixed by Python's grammar, so Ace sidesteps the expression problem almost entirely (and thus inherits broad tool support, as described earlier). The key insight is that leaving the forms fixed does not mean the semantics must also be fixed. Instead of taking a syntax-directed view of extensibility, we take a type-directed view: users cannot add new forms but they can add new type constructors, which we empower with  more control over the semantics of existing forms (e.g. literals, as discussed above, but also nearly every other form, as we will now discuss) in a controlled manner.  %In Ace (but not in the theory in Sec. \ref{theory}), users can also repurpose existing forms flexibly within a delimited scope. 
The key features that characterize active typechecking and translation are: \begin{itemize}
\item a \emph{dispatch protocol} that delegates responsibility for typechecking and translating a term to:
  \begin{itemize}
  \item a type or type constructor extracted from a subterm
  \item a per-function base semantics (or simply \emph{base}) for forms for which this is not possible  (variables, literals without ascriptions and most statements)
  \end{itemize}
\item a mechanism for allowing the user to define new types, type constructors and bases from within the language (in particular, \emph{from within the type-level language}) rather than fixing them ahead of time
\end{itemize}
\begin{codelisting}
\begin{lstlisting}
from listing1 import py, A, record
@py
def example(a):
  {a : A}
  x = a.amount
  x = x + x
  return {
    y: {amount: x} (A),
    remark: "Creating an anonymous record"} (record)
\end{lstlisting}
\caption{[\texttt{listing6.py}] The example detailed in Sec. \ref{att}.}
\label{att-example}
\end{codelisting}

To see how this works, let us trace through how Ace processes the simple example in Listing \ref{att-example}.

\subsection{Base Decorators}
Decorators in Python (line 2) are syntactic sugar: we could equivalently have omitted line 3 and inserted the top-level statement \verb|example = py(example)| on the line after the function definition. Note, however, that \verb|py| is not a function but an instance of a class, \verb|examples.py.PyBase|, that inherits from \verb|ace.Base|. It can be used as a decorator because \verb|ace.Base| overloads the call operator, shown in Listing \ref{base}. The \verb|_process| function (not shown) operates as follows:
\begin{enumerate}
\item The Python standard library is used to extract an abstract syntax tree (\verb|AST|) from the function.
\item The closure of the function, together with the globals in the module it is defined in, are extracted to reify its static environment.
\item The argument annotations are processed. If provided in the body, as in our example, \verb|_process| checks that the argument names are spelled correctly and evaluates the type-level expressions (e.g. \verb|A| above) in the static environment to produce a mapping of argument names to types, called the argument signature. If Python 3's annotations were used, this is not necessary.
%\item Whether the function is generic or not is determined.  We defer discussion of generic functions until Sec. \ref{targets}.
\end{enumerate}
By the end of Listing \ref{att-example}, \verb|example| is thus an instance of \verb|ace.TypedFn|. It could  have been constructed directly by calling the constructor on line 2 of Listing \ref{base} (this would be inconvenient, but could be useful for metaprogramming).

\subsection{Checking Typed Functions}
When the Ace compiler begins typechecking a \verb|TypedFn| (when asked to by \verb|acec|, or when it is  invoked interactively, not discussed in this paper), it proceeds as follows:
\begin{enumerate}
\item An \verb|ace.Context| is constructed with a reference to the function (Listing \ref{context}, lines 2-3).
\item The base is asked to initialize the context. This can be seen on lines 2-4 of Listing \ref{pybase}, where the \verb|PyBase| base adds an attribute tracking local bindings (initially only the arguments) and the return type, which will be synthesized but initially is not known. 
\item For each statement in the body (after the documentation string and type annotation), the compiler delegates to either the base, or of a type synthesized from a subterm, by calling a method named \verb|check_|$F$, where $F$ is the form of the statement (derived directly from Python's \verb|ast| package \cite{python}). These methods are responsible for checking that the statement is well-typed (raising an \verb|ace.TypeError| if not) and having any needed effect on the context. 
\item The base synthesizes a type for the function as a whole via the \verb|syn_FunctionDef_outer| method. Here, an \verb|arrow| type is synthesized (lines 26-29) based on the argument signature and the synthesized return type.
\end{enumerate}
The assignment statements on lines \ref{att-example}.5\footnote{In this section, we will need to refer to code in Listings \ref{att-example}, \ref{pybase}, \ref{context} and \ref{record}, so we adopt this syntactic convention to refer to line numbers for concision.} and \ref{att-example}.6 are checked by the base method \verb|check_Assign_Name| shown on lines \ref{pybase}.6-\ref{pybase}.14 and the return statement on line \ref{att-example}.7 is checked by the base's \verb|check_Return| on lines \ref{pybase}.19-\ref{pybase}.23. Both work similarly: if this is the first assignment to a particular name, or the first return statement seen, then the value must be able to synthesize a type in the current context. Otherwise, it is analyzed against the previously synthesized type.

\subsection{Bidirectional Typechecking}
\begin{codelisting}
\begin{lstlisting}
class TypedFn(object):
  def __init__(self, base, ast, static_env, arg_sig):
    # ...
  # ...
class Base(object):
  def __call__(self, f):
    (ast, static_env, arg_sig) = _process(f)
    return TypedFn(self, ast, static_env, arg_sig)
  # ...
\end{lstlisting}
\caption{A portion of the \texttt{ace} core showing how a base can be used as a decorator to construct a typed function.}
\label{base}
\end{codelisting}

\begin{codelisting}
\begin{lstlisting}
class PyBase(ace.Base):
  def init_ctx(self, ctx):
    ctx.locals = dict(ctx.fn.arg_sig)
    ctx.return_t = None

  def check_Assign_Name(self, ctx, s):
    x, e = s.target.id, s.value
    if x in ctx.locals:
      ctx.ana(e, ctx.locals[x])
    else:
      ty = ctx.syn(e)
      ctx.locals[x] = ty

  def trans_Assign_Name(self, ctx, target, s):
    return target.direct_translation(ctx, s)

  def check_Return(self, ctx, s):
    if ctx.return_t == None:
      ctx.return_t = ctx.syn(s.value)
    else:
      ctx.ana(s.value, ctx.return_t)

  def trans_Return(self, ctx, target, s):
    return target.direct_translation(ctx, s)

  def syn_FunctionDef_outer(self, ctx, f):
    if ctx.return_t == None:
      ctx.return_t = unit
    return arrow[ctx.fn.arg_sig, ctx.return_t]

  def trans_FunctionDef_outer(self, ctx, target, f):
    if f.name == "__toplevel__":
      return target.Suite(ctx.trans(f.body))
    else:
      return target.direct_translation(f, ctx)

  def syn_Name(self, ctx, e):
    x = e.id
    if x in ctx.locals:
      return ctx.locals[x]
    elif x in ctx.fn.static_env: # (Sec. 5)
      return self.syn_lifted(x)
    else:
      raise ace.TypeError("...var not bound...", e)

  def trans_Name(self, ctx, target, e):
    if e.id in ctx.locals:
      return target.direct_translation(ctx, e)
    else: # (Sec. 5)
      return self.trans_lifted(ctx, target, e)

  default_Dict_asc = default_Str_asc = dyn
  # ...
  def init_target(ctx): return PyTarget()   
py = PyBase()
\end{lstlisting}
\caption{A portion of the base used in our examples thus far, defined in the \texttt{examples.py} package.}
\label{pybase}
\end{codelisting}

\begin{codelisting}
\begin{lstlisting}
class Context(object):
  def __init__(self, fn):
    self.fn = fn

>>>>

\title{A Type System for String Sanitation\\Implemented Inside a Python}

%\authorinfo{~}{~}{~}
\authorinfo{Nathan Fulton\and Cyrus Omar\and Jonathan Aldrich}
           {School of Computer Science\\
            Carnegie Mellon University}
           {\{nfulton, comar, aldrich\}@cs.cmu.edu}

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
ABSTRACT HERE
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}\label{intro}
INTRO HERE

\section{A Type System for String Sanitation}

\section{Implementation Inside a Python}
\subsection{Background: Ace}
TODO: Make a TR out of the OOPSLA submission.
\subsection{Explicit Conversions}
\subsection{Adding Subtyping to Ace}
\subsection{Theory}

\section{Related Work}

\section{Discussion}
